Certified Symbolic Finite Transducers: Formalization and Applications to String Manipulation

### 2025-08-26 Paper/application.tex
Added new section
motivation seciton

\section{Applications and Evaluation}

We evaluate our formalization and implementation of the SFT product operation in two key application domains: string sanitization and string replacement operations in string solving. These applications demonstrate the practical utility of our certified SFT product operation in real-world scenarios.
\subsection{Sanitizer}

@@

  \path[->] 
    % Transition for safe characters
    (q0) edge[bend left=20] node[above,midway=1] {$\texttt{safe\_chars}, \texttt{id}$} (q1)
    % Transition for unsafe characters  
    (q0) edge[bend right=20] node[left,midway] {$\neg\texttt{safe\_chars}, \texttt{encode}$} (q2)
    % Epsilon transitions back to q0
    (q1) edge[bend left=30] node[below,midway] {$\varepsilon, \lambda\varepsilon$} (q0)
    (q2) edge[bend right=20] node[right,midway] {$\varepsilon, \lambda\varepsilon$} (q0);
\end{tikzpicture}
\caption{SFT representation of the HTML encoding sanitizer function.}
\label{fig:html-sanitizer-sft}
\end{figure}

@@

Beyond performance evaluation, our SFT formalization enables formal verification of string sanitizer correctness. A santizer is used to transform user inputs into safe outputs, ensuring that the output does not contain any unsafe characters. However, santizers just like programs, which are eror-prone. Manual reviews of sanitizers are often insufficient to guarantee correctness.

Based on our formalization, we can verify sanitizers by modeling the sanitizer as an SFT and the user input as an SFA. We can also model potential attacks as SFAs that accept unsafe strings. Formally speaking, Let $\mathcal{T}$ be an SFT, $\mathcal{I}$ is the possible user inputs modeled by an SFA, and $\mathcal{A}$ is the attack SFA that accepts unsafe strings. The santizer correctness checking against the attack is to check whether the intersection of the product $\mathcal{T} \times \mathcal{I}$ and the attack SFA $\mathcal{A}$ is empty, i.e., $\mathcal{T} \times \mathcal{I} \cap \mathcal{A} = \emptyset$.

Moreover, santizers are usually used in combination with other sanitizers to ensure comprehensive safety. For example, the HTML encoding sanitizer can be combined with a string trimming operation to remove leading and trailing whitespace characters. For this kind of sanitizer composition, it can be modeled by our SFT product as well. Let $\mathcal{T}_1$ and $\mathcal{T}_2$ correspond to two sanitizers, and $\mathcal{T}_1$ is applied before $\mathcal{T}_2$. What we need to check is $\mathcal{T}_2 \times (\mathcal{T}_1 \times \mathcal{I}) \cap \mathcal{A} = \emptyset$.

@@

Table \ref{tab:sanitizer_attack} summarizes the verification results for various sanitizers against different attack models. The first column lists the attack models, the second column indicates whether the sanitizer is classified as safe or unsafe with respect to the attack, and the third column reports the size of the transducers used in the verification, given as [number of states/number of transitions]. For composed SFTs, the size reflects the total number of states and transitions across all components.

@@

\begin{enumerate}
  \item Construct an SFA recognizing the pattern $\texttt{/[0-9]+/}$ and transform it into an SFT by augmenting each transition with output function $f = \lambda x.~\texttt{None}$ that produces the empty string.
  \item Construct an SFA accepting the replacement string "\texttt{NUM}". Since constant strings are specialized regular expressions, we build their SFA representation and convert it to an SFT by adding $\varepsilon$-transitions with output functions that emit the replacement characters sequentially.
  \item Compose the two SFTs by connecting the accepting state of the pattern-matching SFT to the initial state of the replacement-generating SFT via $\varepsilon$-transitions.
\end{enumerate}

@@

    \path[->] 
    (q0'') edge node {\texttt{None}, \texttt{(g 1)}} (q1'')
    (q1'') edge [bend left=5] node[left] {\texttt{None}, \texttt{(g 2)}} (q2'')
    (q2'') edge node {\texttt{None}, \texttt{(g 3)}} (q3'')
    (q1') edge [bend right=30] node {\texttt{None}, \texttt{f}} (q0'');
  \end{tikzpicture}
  \caption{The SFT for the replacement operation \\ $\texttt{replace}(s, \texttt{/[0-9]+/}, \text{``NUM''})$}
  \label{fig-rearranged-automata}
  \end{figure}

@@

\end{tikzpicture}
\caption{The modeling of replace all operation.}
\label{fig:replace-all}
\end{figure}

@@

We evaluate CertiStrR using benchmarks from SMT-LIB 2024 \cite{smtlib_benchmarks}, focusing on the QF\_S and QF\_SLIA logic fragments. The benchmarks are categorized into two groups: (1) singleton matching and (2) replace-all matching.

@@

An "Inconclusive" result indicates that the solver cannot determine satisfiability due to the inherent incompleteness of the string solver's algorithm, not timeout (all tests completed within one minute). This occurs when the solver cannot determine whether string variables can be assigned values that satisfy all constraints.

@@

Notably, PCP and RNA benchmarks cannot be solved by most state-of-the-art string solvers, including CVC4 \cite{cvc4} and Z3 \cite{Z3-str3}, yet CertiStrR solves them efficiently. Among existing solvers, only Ostrich \cite{pacmpl/ChenFHHHKLRW22} can handle these benchmarks.

>>>>

\section{Applications and Evaluation}
We evaluate our formalization and implementation of the SFT product operation across two key application domains: string sanitization and string replacement in string solving. These case studies highlight the practical value of our certified SFT product operation in real-world scenarios.
\subsection{Sanitizer}

@@

  \path[->] 
    % Transition for safe characters
    (q0) edge[bend left=20] node[above,midway=1] {$\texttt{safe\_chars}, \texttt{id}$} (q1)
    % Transition for unsafe characters  
    (q0) edge[bend right=20] node[left,midway] {$\neg\texttt{safe\_chars}, \texttt{encode}$} (q2)
    % Epsilon transitions back to q0
    (q1) edge[bend left=30] node[below,midway] {$\varepsilon, \lambda\varepsilon$} (q0)
    (q2) edge[bend right=20] node[right,midway] {$\varepsilon, \lambda\varepsilon$} (q0);
\end{tikzpicture}
\caption{SFT representation of the HTML encoding sanitizer function.}
\label{fig:html-sanitizer-sft1}
\end{figure}

@@

Beyond performance evaluation, our SFT formalization also supports the formal verification of string sanitizer correctness. A sanitizer transforms user inputs into safe outputs, ensuring that the result contains no unsafe characters. However, sanitizers—like any other programs—are error-prone. Manual reviews or testing alone are often insufficient to guarantee their correctness.

Based on our formalization, we can verify sanitizers by modeling the sanitizer as an SFT and the user input as an SFA. We can also model potential attacks as SFAs that accept unsafe strings. Formally speaking, Let $\mathcal{T}$ be an SFT, $\mathcal{I}$ is the possible user inputs modeled by an SFA, and $\mathcal{A}$ is the attack SFA that accepts unsafe strings. The santizer correctness checking against the attack is to check whether the intersection of the product $\mathcal{T} \times \mathcal{I}$ and the attack SFA $\mathcal{A}$ is empty, i.e., $(\mathcal{T} \times \mathcal{I}) \cap \mathcal{A} = \emptyset$.

Moreover, santizers are usually used in combination with other sanitizers to ensure comprehensive safety. For example, the HTML encoding sanitizer can be combined with a string trimming operation to remove leading and trailing whitespace characters. For this kind of sanitizer composition, it can be modeled by our SFT product as well. Let $\mathcal{T}_1$ and $\mathcal{T}_2$ correspond to two sanitizers, and $\mathcal{T}_1$ is applied before $\mathcal{T}_2$. What we need to check is $(\mathcal{T}_2 \times (\mathcal{T}_1 \times \mathcal{I})) \cap \mathcal{A} = \emptyset$.

@@

Table \ref{tab:sanitizer_attack} summarizes the verification results for various sanitizers against different attack models. The first column lists the attack models, the second column indicates whether the sanitizer is classified as safe or unsafe with respect to the attack, and the third column reports the size of the transducers used in the verification, given as [number of states/number of transitions]. For composed SFTs, the size reflects the total number of states and transitions across all SFTs.

@@

\begin{enumerate}
  \item Construct an SFA recognizing the pattern $\texttt{/[0-9]+/}$ and transform it into an SFT by augmenting each transition with output function "$f = \lambda x.~\texttt{None}$" that produces the empty string.
  \item Construct an SFA accepting the replacement string "\texttt{NUM}". Since constant strings are specialized regular expressions, we build their SFA representation and convert it to an SFT by adding $\varepsilon$-transitions with output functions that emit the replacement characters sequentially.
  \item Compose the two SFTs by connecting the accepting state of the pattern-matching SFT to the initial state of the replacement-generating SFT via $\varepsilon$-transitions.
\end{enumerate}

@@

    \path[->] 
    (q0'') edge node {\texttt{None}, \texttt{(g 1)}} (q1'')
    (q1'') edge [bend left=5] node[left] {\texttt{None}, \texttt{(g 2)}} (q2'')
    (q2'') edge node {\texttt{None}, \texttt{(g 3)}} (q3'')
    (q1') edge [bend right=30] node {\texttt{None}, \texttt{f}} (q0'');
  \end{tikzpicture}
  \caption{The SFT for the replacement operation \\ $\texttt{replace}(s, \texttt{/[0-9]+/}, \text{``NUM''})$}
  \label{fig-rearranged-automata1}
  \end{figure}

@@

\end{tikzpicture}
\caption{The modeling of replace all operation.}
\label{fig:replace-all1}
\end{figure}

@@

We evaluate CertiStrR using benchmarks from SMT-LIB 2025 \cite{smtlib_benchmarks}, focusing on the QF\_S and QF\_SLIA logic fragments. The benchmarks are categorized into two groups: (1) singleton matching and (2) replace-all matching.

@@

An "Inconclusive" result indicates that the solver cannot determine satisfiability due to the inherent incompleteness of the string solver's algorithm, not timeout (all tests completed within one minute). This occurs when the solver cannot determine whether string variables need to be assigned different values that satisfy all constraints.

@@

Notably, PCP and RNA benchmarks cannot be solved by most state-of-the-art string
solvers, including CVC4 \cite{cvc4} and Z3 \cite{Z3-str3}, yet CertiStrR solves
them efficiently. Among existing solvers, only OSTRICH \cite{pacmpl/ChenFHHHKLRW22} can handle these benchmarks.

### 2025-08-26 Paper/formalization.tex
finish application and evaluation part
motivation seciton

\section{The Isabelle/HOL Formalization of SFTs}

We present the basic architecture of our SFT formalization in Isabelle/HOL, which consists of three distinct layers.

@@

The \textbf{implementation layer} refines the abstract transition labels to a locale where a boolean algebra is defined as type variable \texttt{'b} together with a collection of operations and assumptions. State sets are refined using Refine\_Monadic \cite{Refine_Monadic-AFP}, a framework in Isabelle/HOL for data refinement that automatically transforms general sets into efficient data structures such as hashmaps and red-black trees.

>>>>

\section{The Isabelle/HOL Formalization of SFTs}
\label{sec:formalization}

In this section, we present our formalization of SFTs. We begin by outlining the basic architecture of our Isabelle/HOL formalization, which is organized into three distinct layers.

@@

The \textbf{implementation layer} refines the abstract transition labels to a locale where a boolean algebra is defined as type variable $'\texttt{b}$ together with a collection of operations and assumptions. State sets are refined using Refine\_Monadic \cite{Refine_Monadic-AFP}, a framework in Isabelle/HOL for data refinement that automatically transforms general sets into efficient data structures such as hashmaps and red-black trees.
